---
title: "The interplay between glucogenic amino acids and carbohydrate metabolisms in regulating the clinical fate of infected preterm newborns"
subtitle: "Plasma metabolomics - EDA - PCA"
author: 
- "Person in Charge: Ziyuan WU, Nguyen Tran Nam Tien, Nguyen Thi Hai Yen"
- "Supervisor: Duc Ninh Nguyen, Nguyen Phuoc Long"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
   includes:
    in_header: "wrap-code.tex"
   keep_tex: true
   number_sections: true
   toc: true
---

<style type="text/css">

body{
  font-family: Arial;
  font-size: 12pt;
}
code.r{
  font-family: FiraCode;
  font-size: 14px;
}
pre {
  font-size: 13px
}
</style>

```{r setup, include=FALSE, fig.align='center',tidy=TRUE}
# SET UP CODE CHUNK FOR REPORTING PURPOSE

knitr::opts_chunk$set(echo = TRUE)

# Remove all R objects in the environment first
rm(list=ls())

# Wrap output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
    # this hook is used only when the linewidth option is not NULL
    if (!is.null(n <- options$linewidth)) {
        x = xfun::split_lines(x)
        # any lines wider than n should be wrapped
        if (any(nchar(x) > n))
            x = strwrap(x, width = n)
        x = paste(x, collapse = "\n")
    }
    hook_output(x, options)
})

# Table reporting
show_table = function(df, show_rowname = TRUE){
  library(DT) 
  
  df %>% datatable(
    class = 'cell-border stripe', filter = 'top',
    extensions = list('Scroller' = NULL, 'Buttons' = NULL),
    options = list(
      deferRender = TRUE,
      scrollY = 300,
      scrollX = TRUE,
      dom = 'Bfrtip', buttons = c('colvis', 'excel', 'csv', 'pdf', 'copy'),
       initComplete = JS(
        "function(settings, json) {",
        "$(this.api().table().body()).css({'font-size': '80%'});",
        "}")
    ),
    rownames = show_rowname
  )
}

# Another one of table reporting
knitr_table = function(df, caption = ""){
  library(kableExtra)
  
  if (caption == "") {
    df %>% kbl() %>% kable_paper(bootstrap_options = "striped", full_width = F, html_font = "Arial")
  } else {
    df %>% kbl(caption = caption) %>% kable_paper(bootstrap_options = "striped", full_width = F, html_font = "Arial")
  }
}

# Insert image
insert_image = function(dir, image_name){
  knitr::include_graphics(
    base::paste(
      dir, "/",
      image_name,
      sep = ""
    )
  )
}

# Define time to export the data
exported_time = "240205"
```

# Call packages
```{r warning=FALSE,message=FALSE}
# Data wrangling
library(tidyverse)
library(openxlsx)

# Metabolomics data analysis
library(MetaboAnalystR)
library(quotidieR)
library(ggpubr)

`%!in%` = Negate(`%in%`)
```

# Note
- Check the `MetaboAnalystR` version (see `sessionInfo()` section). However, it should be noted that even with the same `MetaboAnalystR` and `R` versions, the code might not work. This might be due to the day of installation. Thus, when the code does not work, please find 4 below lines in this code and remove them, then use the next line (have a comment with the note: new updated function) for these lines.
  + `mSet <- FilterVariable(mSet, "iqr", IQR_cut_off, "T", QC_RSD, F)` (3 lines from PCA, PLS_DA, and linear model function)
  + `mSet <- FilterVariable(mSet, "iqr", IQR_cut_off, "F", QC_RSD, F)` (in linear model function)
  
# Get group label
- Code to obtain groups levels
```{r}
Get_group_level = function(dir, filename){
  
  library(MetaboAnalystR)
  library(tidyverse)
  #Set working directory
  setwd(dir)
  file_path = paste(dir, filename, sep = "/")
  
  mSet <- InitDataObjects("pktable", "stat", FALSE)
  mSet <- Read.TextData(
    mSet, 
    file_path, 
    "colu", "disc"
  )
  
  mSet$dataSet$cls %>% unique() -> group_level
  
  return(group_level)
}
```

- Get group label. 
```{r}
feature.nm.vec <- c("")
smpl.nm.vec <- c("")
# Note: this need in order levels when read the data
group_order = Get_group_level(
  dir = "../2MetaboAnalyst",
  filename = "2Merged_annotated_POS_NEG_Exp2SE.csv"
)
grp.nm.vec <- group_order
```

- Should check our group label.
```{r}
group_order
```

# Data Exploration - PCA
## Define function
```{r}
EDA_PCA = function(dir, filename, 
                   time = as.character(format(Sys.time(), "%Y%m%d")), 
                   QC_RSD = 25, IQR_cut_off = 0,
                   norm_method = "MedianNorm", trans_method = "LogNorm", scale_method = "ParetoNorm",
                   show_label = 0){
  
  library(MetaboAnalystR)
  #Set working directory
  setwd(dir)
  file_path = paste(dir, filename, sep = "/")
  
  # Initiate object
  mSet <- InitDataObjects("pktable", "stat", FALSE)
  # Read data
  mSet <- Read.TextData(
    mSet, 
    file_path, 
    "colu", "disc"
  )
  
  # Check data set as well as NA
  mSet <- SanityCheckData(mSet)
  
  # Data Treatment
  ## NA values remove
  mSet <- RemoveMissingPercent(mSet, percent = 0.5)
  ## NA values imputation
  mSet <- ImputeMissingVar(mSet, method = "knn_var")
  
  mSet <- SanityCheckData(mSet)

  ## QCs repeatability check by threshold of 25%
  mSet <- FilterVariable(mSet, "iqr", IQR_cut_off, "T", QC_RSD, F)   # old version, laptop
  # mSet <- FilterVariable(mSetObj = mSet, qc.filter = "T", rsd = QC_RSD, var.filter = "iqr", var.cutoff = IQR_cut_off)  # new updated function, workstation
  
  ## Decide to retain or remove any groups, note: remove/retain need to be done outside the function
  mSet <- GetGroupNames(mSet, "")
  feature.nm.vec <- c("")
  smpl.nm.vec <- c("")
  grp.nm.vec <- c("")
  mSet <- UpdateData(mSet)
  
  ## Normalize for PCA
  mSet <- PreparePrenormData(mSet)
  mSet <- Normalization(mSet, norm_method, trans_method, scale_method, ratio = FALSE, ratioNum = 20)
  mSet <- PlotNormSummary(mSet, "norm_0_", "png", 300, width=NA)
  mSet <- PlotSampleNormSummary(mSet, "snorm_0_", "png", 300, width=NA)

  # PCA 
  mSet <- PCA.Anal(mSet)
  mSet <- PlotPCAPairSummary(mSet, "pca_pair_0_", "png", 300, width=NA, 5)
  mSet <- PlotPCAScree(mSet, "pca_scree_0_", "png", 300, width=NA, 5)
  mSet <- PlotPCA2DScore(mSet, "pca_score2d_0_", "png", 300, width=NA, 1,2,0.95,show_label,0)
  mSet <- PlotPCA3DScore(mSet, "pca_score3d_0_", "json", 1,2,3)
  mSet <- PlotPCA3DScore(mSet, "pca_score3d_0_", "png", 1,2,3)
  mSet <- PlotPCALoading(mSet, "pca_loading_0_", "png", 300, width=NA, 1,2)
  mSet <- PlotPCABiplot(mSet, "pca_biplot_0_", "png", 300, width=NA, 1,2)
  mSet <- PlotPCA3DLoading(mSet, "pca_loading3d_0_", "json", 1,2,3)
  
  # Save data sets
  mSet <- SaveTransformedData(mSet)
}
```

## PCA of interested groups
```{r linewidth=100}
feature.nm.vec <- c("")
smpl.nm.vec <- c("")
# Retain 
grp.nm.vec <- group_order[c(2:3)]

EDA_PCA(
  dir = "../2MetaboAnalyst",
  filename = "2Merged_annotated_POS_NEG_Exp2SE.csv",
  QC_RSD = 25, IQR_cut_off = 0,
  norm_method = "MedianNorm", trans_method = "LogNorm", scale_method = "ParetoNorm"
)
```

```{r echo=FALSE, fig.height=4.6, fig.width=4.6, dpi=600}
insert_image(
  dir = "../2MetaboAnalyst",
  image_name = "pca_score2d_0_dpi300.png"
)
```

## Manual visualization - PCA
```{r warning=FALSE, fig.height = 5, fig.width = 10}
# From normalized and scaled data
dir_pca = "../2MetaboAnalyst"

input_pca = read.csv(
  paste(dir_pca, "data_normalized.csv", sep = "/"),
  row.names = 1
)

input_pca %>% 
  dplyr::slice(-1) %>%  
  mutate_if(is.character, as.numeric) %>% 
  #select_if(function(col) !all(col == 0))
  filter(rowSums(. == 0) != ncol(.)) %>% #remove 0 values across samples
  t() %>% 
  prcomp(scale. = FALSE) -> pca

summary_pca = summary(pca)
summary_pca$importance %>% as.data.frame() -> summary_pca

### Read metadata
metadata = read.csv(
  paste(dir_pca, "0GAA_metadata_Exp2SE.csv", sep = "/")
) %>% mutate(Gender = toupper(Gender))

# Visualization
Exp2SE <- as_tibble(pca$x[,1:2]) %>%
                 mutate(Sample = names(input_pca)) %>% 
                 inner_join(metadata,
                            by = "Sample") %>% 
                 mutate(Group = factor(Group, levels = c("GLUSE", "GAASE"),labels = c("SE-sGLU","SE-rGLU-GAAs"))) %>% 
                 ggscatter(x = "PC1", y = "PC2", color = "Group", size = 1.5, star.plot = TRUE, star.plot.lwd = 0.2)+
                 scale_color_manual(values = c("#009193", "#EB6F69"))+
                 stat_ellipse(aes(x=PC1,y=PC2,fill=Group),geom="polygon",level=0.95, alpha=0.15)+
                 scale_fill_manual(values = c("#009193", "#EB6F69"))+
                 labs(x = paste("PC1 (", round(summary_pca$PC1[2]*100, 1), " %)", sep = ""), 
                                     y = paste("PC2 (", round(summary_pca$PC2[2]*100, 1), " %)", sep = ""))+
                 theme_bw() +
                 theme(panel.border = element_rect(colour = "black", fill = NA, linewidth= 0.5),    
                       panel.grid.minor = element_blank(),
                       panel.grid.major = element_blank())+
                 theme(panel.border = element_rect(colour = "black", fill = NA, linewidth= 0.5),    
                       panel.grid.minor = element_blank(),
                       panel.grid.major = element_blank())+
                 theme(legend.text=element_text(size=16))+
                 theme(legend.title=element_text(size=18))+
                 theme(strip.text.x = element_text(size = 16))+
                 theme(strip.text.y = element_text(size = 16))+
                 theme(axis.title.y=element_text(size=18,color="black"),
                       axis.title.x=element_text(size=18,color="black"),
                       axis.text.y=element_text(size=16,color="black"),
                       axis.text.x=element_text(size=16,color="black"))
                               
ggsave(Exp2SE,filename = "../2MetaboAnalyst/PCAplot_Exp2SE.png",width = 7.5,height = 5,device = "png")
```

# sessionInfo()
```{r}
sessionInfo()
```

