---
title: "The interplay between glucogenic amino acids and carbohydrate metabolisms in regulating the clinical fate of infected preterm newborns"
subtitle: "Plasma metabolomics - Statistical analysis"
author: 
- "Person in Charge: Ziyuan WU, Nguyen Tran Nam Tien, Nguyen Thi Hai Yen"
- "Supervisor: Duc Ninh Nguyen, Nguyen Phuoc Long"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
   includes:
    in_header: "wrap-code.tex"
   keep_tex: true
   number_sections: true
   toc: true
---

<style type="text/css">

body{
  font-family: Arial;
  font-size: 12pt;
}
code.r{
  font-family: FiraCode;
  font-size: 14px;
}
pre {
  font-size: 13px
}
</style>

```{r setup, include=FALSE, fig.align='center',tidy=TRUE}
# SET UP CODE CHUNK FOR REPORTING PURPOSE

knitr::opts_chunk$set(echo = TRUE)

# Remove all R objects in the environment first
rm(list=ls())

# Wrap output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
    # this hook is used only when the linewidth option is not NULL
    if (!is.null(n <- options$linewidth)) {
        x = xfun::split_lines(x)
        # any lines wider than n should be wrapped
        if (any(nchar(x) > n))
            x = strwrap(x, width = n)
        x = paste(x, collapse = "\n")
    }
    hook_output(x, options)
})

# Table reporting
show_table = function(df, show_rowname = TRUE){
  library(DT) 
  
  df %>% datatable(
    class = 'cell-border stripe', filter = 'top',
    extensions = list('Scroller' = NULL, 'Buttons' = NULL),
    options = list(
      deferRender = TRUE,
      scrollY = 300,
      scrollX = TRUE,
      dom = 'Bfrtip', buttons = c('colvis', 'excel', 'csv', 'pdf', 'copy'),
       initComplete = JS(
        "function(settings, json) {",
        "$(this.api().table().body()).css({'font-size': '80%'});",
        "}")
    ),
    rownames = show_rowname
  )
}

# Another one of table reporting
knitr_table = function(df, caption = ""){
  library(kableExtra)
  
  if (caption == "") {
    df %>% kbl() %>% kable_paper(bootstrap_options = "striped", full_width = F, html_font = "Arial")
  } else {
    df %>% kbl(caption = caption) %>% kable_paper(bootstrap_options = "striped", full_width = F, html_font = "Arial")
  }
}

# Insert image
insert_image = function(dir, image_name){
  knitr::include_graphics(
    base::paste(
      dir, "/",
      image_name,
      sep = ""
    )
  )
}

# Write `csv` files
write_csv_for_target_name = function(df, dir, time = as.character(format(Sys.time(), "%Y%m%d")),
                                     name_aim, mode, rowname = FALSE){
  name = paste(dir, paste(paste(time, name_aim, mode, sep = "_"), "csv", sep = "."), sep = "/")
  
  write.csv(df, name, row.names = rowname)
}

# Define time to export the data
exported_time = "240512"
```

# Call packages
```{r warning=FALSE,message=FALSE}
# Data wrangling
library(tidyverse)
library(openxlsx)

# Metabolomics data analysis
library(MetaboAnalystR)

# Statistical analysis
library(lme4)
library(multcomp)
library(multtest)

`%!in%` = Negate(`%in%`)
```

# Note
- Check the `MetaboAnalystR` version (see `sessionInfo()` section). However, it should be noted that even with the same `MetaboAnalystR` and `R` versions, the code might not work. This might be due to the day of installation. Thus, when the code does not work, please find 4 below lines in this code and remove them, then use the next line (have a comment with the note: new updated function) for these lines.
  + `mSet <- FilterVariable(mSet, "iqr", IQR_cut_off, "T", QC_RSD, F)` (3 lines from PCA, PLS_DA, and linear model function)
  + `mSet <- FilterVariable(mSet, "iqr", IQR_cut_off, "F", QC_RSD, F)` (in linear model function)
  
# Get group label
- Code to obtain groups levels.
```{r}
Get_group_level = function(dir, filename){
  
  library(MetaboAnalystR)
  library(tidyverse)
  #Set working directory
  setwd(dir)
  file_path = paste(dir, filename, sep = "/")
  
  mSet <- InitDataObjects("pktable", "stat", FALSE)
  mSet <- Read.TextData(
    mSet, 
    file_path, 
    "colu", "disc"
  )
  
  mSet$dataSet$cls %>% unique() -> group_level
  
  return(group_level)
}
```

- Get group label. 
```{r}
feature.nm.vec <- c("")
smpl.nm.vec <- c("")
# Note: this need in order levels when read the data
group_order = Get_group_level(
  dir = "../2limma",
  filename = "2Merged_annotated_POS_NEG_Exp3SE.csv"
)
grp.nm.vec <- group_order
```

- Should check our group label.
```{r}
group_order
```

# Statistical Analyses - Linear model
## Linear model
- Function to perform linear model
```{r}
STAT_linear_model <- function(dir, filename_data, filename_metadata, 
                              primary_metadata,
                              group_ref,
                              QC_RSD = 25, IQR_cut_off = 0,
                              norm_method = "MedianNorm", trans_method = "LogNorm", scale_method = "NULL",
                              p_val = 0.05, F_C, FDR 
                              ) {
  library(MetaboAnalystR)
  library(tidyverse)
  library(quotidieR)
  
  # Prepare data set for STAT_DEMs: NA_imputted_QC_filltered
  setwd(dir)
  file_path = paste(dir, filename_data, sep = "/")
  
  # Initiate object
  mSet <- InitDataObjects("pktable", "stat", FALSE)
  # Read data
  mSet <- Read.TextData(
    mSet, 
    file_path, 
    "colu", "disc"
  )
  
  # Check data set and NA values
  mSet <- SanityCheckData(mSet)
  
  # Data Treatment
  ## NA values remove
  mSet <- RemoveMissingPercent(mSet, percent = 0.5)
  ## NA values imputation
  mSet <- ImputeMissingVar(mSet, method = "knn_var")
  
  mSet <- SanityCheckData(mSet)
  
  ## QCs repeatability check by threshold of 25%
  mSet <- FilterVariable(mSet, "iqr", IQR_cut_off, "T", QC_RSD, F)
  #mSet <- FilterVariable(mSetObj = mSet, qc.filter = "T", rsd = QC_RSD, var.filter = "iqr", var.cutoff = IQR_cut_off)  # new updated function
  
  mSet <- SaveTransformedData(mSet)
  
  data_processed = read.csv("data_processed.csv", row.names = 1, header = F)
  data_processed %>% 
    janitor::row_to_names(row_number = 1) %>% 
    t() %>% 
    as.data.frame() %>% 
    filter(Label %in% grp.nm.vec) %>% # retain only interested groups (define outside the function)
    t() %>% 
    as.data.frame() %>% 
    write.csv("data_STAT_DEMs_input.csv")
  
  rm(mSet)
  
  # STAT_DEMs
  setwd(dir)
  mSet <- InitDataObjects("pktable", "mf", paired = FALSE)
  mSet<-SetDesignType(mSet, "multi")
  
  file_path_data = "data_STAT_DEMs_input.csv"
  file_path_metadata = paste(dir, filename_metadata, sep = "/")
  mSet <- Read.TextDataTs(mSet, file_path_data, "colmf")
  mSet <- ReadMetaData(mSet, file_path_metadata) 
  
  # Data Checking and Treatment
  mSet <- SanityCheckData(mSet)
  mSet <- ReplaceMin(mSet)
  mSet <- SanityCheckMeta(mSet, 1)
  mSet <- SetDataTypeOfMeta(mSet)
  mSet <- SanityCheckData(mSet)
  mSet <- FilterVariable(mSet, "iqr", IQR_cut_off, "F", QC_RSD, F)
  #mSet <- FilterVariable(mSetObj = mSet, qc.filter = "F", rsd = QC_RSD, var.filter = "none", var.cutoff = IQR_cut_off) # new updated function
  
  # Data Normalization
  mSet <- PreparePrenormData(mSet)
  mSet <- Normalization(mSet, norm_method, "LogNorm", "NULL", ratio=FALSE, ratioNum=20)
  mSet <- PlotNormSummary(mSet, "norm_0_", "png", 72, width=NA)
  mSet <- PlotSampleNormSummary(mSet, "snorm_0_", "png", 72, width=NA)
  
  # Perform linear model using limma method
  ## Create new object without set cut-off of pvalue to obtain raw p value for all features for functional analysis
  mSet_no_p_val <- CovariateScatter.Anal(
    mSet, paste("covariate_plot", sep = "_"), "png", 
    primary_metadata, group_ref, "NA" , 10, "anova"
  )
  
  ## With raw p_val
  mSet <- CovariateScatter.Anal(
    mSet, paste("covariate_plot", sep = "_"), "png", 
    primary_metadata, group_ref, "NA" , p_val, "anova"
  )
  
  # Output
  ## All features having raw and adjusted p value for functional analysis
  mSet_no_p_val$analSet$cov$sig.mat %>% 
    as.data.frame() %>% rownames_to_column(var = "Features") -> all_features_p_val
  rm(mSet_no_p_val)
  
  ## Sig. features with defined threshold
  mSet$analSet$cov$sig.mat %>% as.data.frame() %>% 
    rownames_to_column(var = "Features") %>% 
    mutate(FC = 10^logFC,
           log2FoldChange = log2(10^logFC)) %>% 
    filter(adj.P.Val < FDR) %>% 
    filter(abs(log2FoldChange) > log2(F_C)) -> sig_features 
  
  ## Num of putative metabolites
  sig_features %>% 
    filter(!grepl("Unknown", Features)) %>% 
    filter(!grepl("w/o", Features)) -> putative_metabolites
  
  putative_metabolites %>% nrow() -> num_of_putative_metabolites
  
  ## Heatmap of DEMs
  ### Create new mSet obj for sure
  mSet_heatmap = mSet
  library(pheatmap)
  mSet <- PlotHeatMap(mSetObj = mSet_heatmap, 
                      var.inx = putative_metabolites$Features,
                      imgName = "heatmap2_1_", format = "png", dpi = 300, font.size = 10, 
                      #width = 0, #for MDAs
                      dataOpt = "norm", scaleOpt = "row", smplDist = "euclidean", clstDist = "ward.D", palette = "bwm",
                      rowV = FALSE)
  
  rm(mSet_heatmap)
  
  ## Manual Plot
  mSetObj = mSet
  both.mat <- mSetObj$analSet$cov.mat
  both.mat <- both.mat[order(-both.mat[,"pval.adj"]),]
  logp_val <- mSetObj$analSet$cov$thresh
  library(ggrepel)
  topFeature <- 5;
  if(nrow(both.mat) < topFeature){
    topFeature <- nrow(both.mat);
  }
  
  cov_plot = ggplot(both.mat, mapping = aes(x = pval.no, y = pval.adj)) +
    geom_point(alpha = 0.5, size = 2) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
    xlab("-log10(P-value): no covariate adjustment") +
    ylab("-log10(P-value): adjusted") +
    coord_fixed() +
    theme_Publication()
  
  # print out results
  print(
    paste("Number of Total Significant Features with F.C", F_C, "and FDR", FDR, "Criteria is", nrow(sig_features))
  )
  print(
    paste("Number of Total Up-Regulated Significant Features with F.C", F_C, "and FDR", FDR, "Criteria is",
          sig_features %>% filter(FC > 1) %>% nrow())
  )
  print(
    paste("Number of Total Down-Regulated Significant Features with F.C", F_C, "and FDR", FDR, "Criteria is",
          sig_features %>% filter(FC < 1) %>% nrow())
  )
  print(paste("Number of putative metabolites is", num_of_putative_metabolites))
  
  # Export data sets
  mSet <- SaveTransformedData(mSet)
  ## all_features. Now, for functional analysis, updated packages have had p-value of all features within covariate_result
  write_csv_for_target_name(
    all_features_p_val,
    dir = dir,
    time = "",
    name_aim = "linear_all",
    mode = ""
  )
  
  ##sig_features_df
  write_csv_for_target_name(
    sig_features,
    dir = dir,
    time = "",
    name_aim = "sig_features",
    mode = ""
  )
  ##covariate_df
  write_csv_for_target_name(
    as.data.frame(mSetObj$analSet$cov.mat),
    dir = dir,
    time = "",
    name_aim = "df_for_covariate_plot",
    mode = ""
  )
  ##putative_metabolites
  write_csv_for_target_name(
    putative_metabolites,
    dir = dir,
    time = "",
    name_aim = "putative_metabolites",
    mode = ""
  )
  
  # object wanna to return
  return(
    list(
      total_up_down_putative_metabolites = c(
        nrow(sig_features),
        sig_features %>% filter(FC > 1) %>% nrow(),  
        sig_features %>% filter(FC < 1) %>% nrow(),
        num_of_putative_metabolites
      ),
      covariate_result = mSet$analSet$cov$sig.mat %>% as.data.frame(),
      sig_features_df = sig_features,
      df_for_covariate_plot = as.data.frame(mSetObj$analSet$cov.mat),
      putative_metabolites = putative_metabolites,
      covariate_plot = cov_plot
    )
  )
}
```

### POS+NEG: merged annotated metablites
```{r}
# Define variables need to be adjusted. This step is necessary to guarantee code run. The purpose is to load `adj.vec` object into the R environment.
adj.vec = c("Birth_weight")

# Retain intersested groups (Survival vs. Non_survival)
feature.nm.vec <- c("")
smpl.nm.vec <- c("")
# Retain 
grp.nm.vec <- group_order[c(2,3)]

Survival_vs_Nonsurvival = STAT_linear_model(
  dir = "../2limma",
  filename_data = "2Merged_annotated_POS_NEG_Exp3SE.csv",
  filename_metadata = "0GAA_metadata_Exp3SE_v2.csv",
  primary_metadata = "Group",
  group_ref = "rGLUSE",
  p_val = 0.05, 
  F_C = 1, 
  FDR = 0.1,
  QC_RSD = 25, IQR_cut_off = 0,
  norm_method = "MedianNorm", trans_method = "LogNorm", scale_method = "NULL"
)
```

#### Heatmap of DEMs
```{r echo=FALSE}
insert_image(
  dir = "../2limma",
  image_name = "heatmap2_1_dpi300.png"
)
```

# sessionInfo()
```{r}
sessionInfo()
```

