---
title: "The interplay between glucogenic amino acids and carbohydrate metabolisms in regulating the clinical fate of infected preterm newborns"
subtitle: "Plasma metabolomics - Statistical analysis"
author: 
- "Person in Charge: Ziyuan WU, Nguyen Tran Nam Tien, Nguyen Thi Hai Yen"
- "Supervisor: Duc Ninh Nguyen, Nguyen Phuoc Long"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
   includes:
    in_header: "wrap-code.tex"
   keep_tex: true
   number_sections: true
   toc: true
---

<style type="text/css">

body{
  font-family: Arial;
  font-size: 12pt;
}
code.r{
  font-family: FiraCode;
  font-size: 14px;
}
pre {
  font-size: 13px
}
</style>

```{r setup, include=FALSE, fig.align='center',tidy=TRUE}
# SET UP CODE CHUNK FOR REPORTING PURPOSE

knitr::opts_chunk$set(echo = TRUE)

# Remove all R objects in the environment first
rm(list=ls())

# Wrap output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
    # this hook is used only when the linewidth option is not NULL
    if (!is.null(n <- options$linewidth)) {
        x = xfun::split_lines(x)
        # any lines wider than n should be wrapped
        if (any(nchar(x) > n))
            x = strwrap(x, width = n)
        x = paste(x, collapse = "\n")
    }
    hook_output(x, options)
})

# Table reporting
show_table = function(df, show_rowname = TRUE){
  library(DT) 
  
  df %>% datatable(
    class = 'cell-border stripe', filter = 'top',
    extensions = list('Scroller' = NULL, 'Buttons' = NULL),
    options = list(
      deferRender = TRUE,
      scrollY = 300,
      scrollX = TRUE,
      dom = 'Bfrtip', buttons = c('colvis', 'excel', 'csv', 'pdf', 'copy'),
       initComplete = JS(
        "function(settings, json) {",
        "$(this.api().table().body()).css({'font-size': '80%'});",
        "}")
    ),
    rownames = show_rowname
  )
}

# Another one of table reporting
knitr_table = function(df, caption = ""){
  library(kableExtra)
  
  if (caption == "") {
    df %>% kbl() %>% kable_paper(bootstrap_options = "striped", full_width = F, html_font = "Arial")
  } else {
    df %>% kbl(caption = caption) %>% kable_paper(bootstrap_options = "striped", full_width = F, html_font = "Arial")
  }
}

# Insert image
insert_image = function(dir, image_name){
  knitr::include_graphics(
    base::paste(
      dir, "/",
      image_name,
      sep = ""
    )
  )
}

# Define time to export the data
exported_time = "240502"
```

# Call packages
```{r warning=FALSE,message=FALSE}
# Data wrangling
library(tidyverse)
library(openxlsx)

# Metabolomics data analysis
library(MetaboAnalystR)

# Statistical analysis
library(lme4)
library(multcomp)
library(multtest)

`%!in%` = Negate(`%in%`)
```

# Note
- Check the `MetaboAnalystR` version (see `sessionInfo()` section). However, it should be noted that even with the same `MetaboAnalystR` and `R` versions, the code might not work. This might be due to the day of installation. Thus, when the code does not work, please find 4 below lines in this code and remove them, then use the next line (have a comment with the note: new updated function) for these lines.
  + `mSet <- FilterVariable(mSet, "iqr", IQR_cut_off, "T", QC_RSD, F)` (3 lines from PCA, PLS_DA, and linear model function)
  + `mSet <- FilterVariable(mSet, "iqr", IQR_cut_off, "F", QC_RSD, F)` (in linear model function)
  
# Get group label
- Code to obtain groups levels.
```{r}
Get_group_level = function(dir, filename){
  
  library(MetaboAnalystR)
  library(tidyverse)
  #Set working directory
  setwd(dir)
  file_path = paste(dir, filename, sep = "/")
  
  mSet <- InitDataObjects("pktable", "stat", FALSE)
  mSet <- Read.TextData(
    mSet, 
    file_path, 
    "colu", "disc"
  )
  
  mSet$dataSet$cls %>% unique() -> group_level
  
  return(group_level)
}
```

- Get group label. 
```{r}
feature.nm.vec <- c("")
smpl.nm.vec <- c("")
# Note: this need in order levels when read the data
group_order = Get_group_level(
  dir = "../2lmer",
  filename = "2Merged_annotated_POS_NEG_Exp2SE.csv"
)
grp.nm.vec <- group_order
```

- Should check our group label.
```{r}
group_order
```

# Data treatment and normalization
## Define function
```{r}
data_treatment_lmer = function(dir, filename, 
                               time = as.character(format(Sys.time(), "%Y%m%d")), 
                               QC_RSD = 25, IQR_cut_off = 0,
                               norm_method = "MedianNorm", trans_method = "LogNorm", scale_method = "ParetoNorm",
                               show_label = 0){
  
  library(MetaboAnalystR)
  #Set working directory
  setwd(dir)
  file_path = paste(dir, filename, sep = "/")
  
  # Initiate object
  mSet <- InitDataObjects("pktable", "stat", FALSE)
  # Read data
  mSet <- Read.TextData(
    mSet, 
    file_path, 
    "colu", "disc"
  )
  
  # Check data set as well as NA
  mSet <- SanityCheckData(mSet)
  
  # Data Treatment
  ## NA values remove
  mSet <- RemoveMissingPercent(mSet, percent = 0.5)
  ## NA values imputation
  mSet <- ImputeMissingVar(mSet, method = "knn_var")
  
  mSet <- SanityCheckData(mSet)

  ## QCs repeatability check by threshold of 25%
  mSet <- FilterVariable(mSet, "iqr", IQR_cut_off, "T", QC_RSD, F)   # old version, laptop
  # mSet <- FilterVariable(mSetObj = mSet, qc.filter = "T", rsd = QC_RSD, var.filter = "iqr", var.cutoff = IQR_cut_off)  # new updated function, workstation
  
  ## Decide to retain or remove any groups, note: remove/retain need to be done outside the function
  ### Remove QCs before normalization
  mSet <- GetGroupNames(mSet, "")
  feature.nm.vec <- c("")
  smpl.nm.vec <- c("")
  grp.nm.vec <- c("")
  mSet <- UpdateData(mSet)
  
  ## Normalize for statistical analysis
  mSet <- PreparePrenormData(mSet)
  mSet <- Normalization(mSet, norm_method, trans_method, scale_method, ratio = FALSE, ratioNum = 20)
  mSet <- PlotNormSummary(mSet, "norm_0_", "png", 300, width=NA)
  mSet <- PlotSampleNormSummary(mSet, "snorm_0_", "png", 300, width=NA)

  # Save data sets
  mSet <- SaveTransformedData(mSet)
}
```

## Treat and normalize data
- Use `data_normalized.csv` as input of statistical analysis.
```{r linewidth=100}
feature.nm.vec <- c("")
smpl.nm.vec <- c("")
# Retain (i.e., wo QC)
grp.nm.vec <- group_order[c(2:3)]

data_treatment_lmer(
  dir = "../2lmer",
  filename = "2Merged_annotated_POS_NEG_Exp2SE.csv",
  QC_RSD = 25, IQR_cut_off = 0,
  norm_method = "MedianNorm", trans_method = "LogNorm", scale_method = "NULL"
)
```

- Standardize data format for `lmer`.
```{r}
# Read normalized data
normalized_data = read.csv(
  "../2lmer/data_normalized.csv",
  row.names = 1
) %>% 
  # remove Label
  dplyr::slice(-1) %>% 
  # Convert to nummeric
  mutate_if(is.character, as.numeric)
  

# Read metadata
metadata = read.csv(
  "../2lmer/0GAA_metadata_Exp2SE.csv"
)

# Merge the 2 above data to Standardize data format for `lmer` [samples in row].
metadata %>% 
  inner_join(
    t(normalized_data) %>% as.data.frame() %>% 
      rownames_to_column(var = "Sample"),
    by = join_by(Sample)
  ) -> lmer_input

# Export
write.csv(
  lmer_input,
  "../2lmer/_lmer_input.csv",
  row.names = FALSE
)
```

# MDAs by `lmer` function
## P-Value calculation
```{r warning=FALSE,message=FALSE}
# Loop for linear mixed model
end.glht <- function(x){
  t1 <- with(lmer_input, data.frame(Group, Gender, Birth_weight, Litter, x)) #extract data to make a temp frame, x means each of metabolite
  colnames (t1) <- c("Group","Gender", "Birth_weight","Litter", "XX") #adding column names
  t2 <- lmer(XX ~ Group + Gender + Birth_weight + (1|Litter), REML= FALSE, data = t1)  
  #compare based on 'Group' levels and extract p.values
  summary(glht(t2, linfct = mcp(Group = "Tukey"), quantile.type = 10))$test$pvalues
}

# Take the location of metabolite in the data.frame
analyte_location = 19:length(lmer_input[1,])

# Apply loop for the statistics of all metabolites using the above location
sbs.tr <- sapply(lmer_input[,analyte_location], FUN = end.glht)
```

## Adjust P-Value
```{r}
# Get rawp vector
rawp = as.vector(sbs.tr)

# by the methods of TSBH with 0.05
sbs.ad_P <- mt.rawp2adjp(rawp, proc = "TSBH", alpha = 0.05)

# Reorder (using index) after adjusting to merge
sbs.ad_P_Reorder <- sbs.ad_P$adjp[order(sbs.ad_P$index),]

# Add metabolites name
sbs.ad_P_Reorder %>% as.data.frame() %>% 
  cbind(data.frame(rawp2 = sbs.tr)) %>% 
  dplyr::rename(padj = TSBH_0.05) -> lmer_result

# Check rawp after merging 
if(identical(lmer_result$rawp, lmer_result$rawp2)) {
  print("Successfully merge data using `cbind()`")
  
  lmer_result %>% dplyr::select(-rawp2) -> lmer_result
} else {
  "Unsuccessfully merge data using `cbind()`. Please check manually again."
}
```

## Number of MDAs
```{r}
# Number of MDAs
FDR_cut_off = 0.1

lmer_result %>% 
  mutate(TSBH_0.1 = ifelse(padj < FDR_cut_off, "o", "x")) -> lmer_result

print(base::paste(
  lmer_result %>% filter(TSBH_0.1 == "o") %>% nrow(), "significant metabolites with FDR cut-off of", FDR_cut_off
))

# Export results
write.csv(
  lmer_result %>% rownames_to_column(var = "Features"),
  "../2lmer/_lmer_result.csv",
  row.names = FALSE
)
```

# sessionInfo()
```{r}
sessionInfo()
```

